\chapter{General}
\label{ch:general}

\section{Purpose}
\paragraph{}This \ac{sop} establishes the operational procedures and operator selection/training requirements necessary to ensure safe flight operations for Project \ac{mars} by personnel within \ac{kach}. This \ac{sop} is designed to comply with all applicable \ac{faa}, \ac{dod}, and Army regulations, as well as \ac{kach} policies.

\section{Description of Operation}
\paragraph{}This \ac{sop} describes piloted and autonomous flight operations, maintenance and safety considerations, and personnel requirements related to the operation of multi-rotor \ac{suas} within the Project \ac{mars} inventory.
\\
\\
\noindent \textbf{Note:} This \ac{sop} must be present in the operational area of use. A copy will also be maintained in the \ac{kach} Project \ac{mars} office.

\section{Applicability}
This \ac{sop} applies to all government and contractor personnel who are actively involved in air or ground operations related to Project \ac{mars}.  Operational procedures described in this \ac{sop} will be used to maintain prudent, safe operating practices and to ensure that appropriate response actions are taken in the event of an emergency.

\section{Proponent and Authority}
\paragraph{Proponent}The proponent for this \ac{sop} is the \ac{kach} Project \ac{oic}.
\\
\paragraph{Exception/Waiver Authority}The proponent may approve written exceptions/waivers to this \ac{sop} that are consistent with controlling law, regulation, and safety policy. The proponent may delegate approval authority in writing to the Project \ac{ncoic}.
\\
\paragraph{Limits to Proponent Authority}Any change that alters \ac{opord}-controlled measures or authorities (for example, \ac{acm}s, altitude blocks, flight corridors/routes/boxes, \ac{tfb} approval lanes, or delegated command authority) requires action by the \ac{usma} G-3 via \ac{frago}.
\\
\paragraph{Recommended Changes}Users submit recommended changes via email to the Project \ac{ncoic} or Project \ac{oic}; the Project \ac{oic} adjudicates and publishes approved updates.
\\
\paragraph{Precendence}In the event of a conflict between this \ac{sop} and controlling law, regulation, or policy, the controlling document takes precedence. The proponent will correct the \ac{sop} at the next revision.
