\chapter{General}
\label{ch:general}

\section{Purpose}
        \subsubsection{}This \ac{sop} establishes the operational procedures and operator selection/training requirements necessary to ensure safe flight operations for \ac{kach} personnel within \projectmars{}. This \ac{sop} is designed to comply with all applicable \ac{faa}, \ac{dod}, and Army regulations, as well as \ac{kach} policies.

        The Army is incorporating \ac{suas} into the  inventory.  But currently there is not a clear description from the Army on how to handle development of \ac{uas} equipment not part of a \ac{por}.  \projectmars{} in cooperation with \ac{usma}'s \ac{rrc} is endeavoring to develop, construct, and test a purpose built \gls{group1} \ac{suas} capable of autonomous \ac{medlog} delivery \ac{bvlos}.  \projectmars{} personnel have developed this \ac{sop} to guide the safe use of all \ac{uas} used in conjunction with this Project.

\section{Description of Operation}
        \subsubsection{}This \ac{sop} describes piloted and autonomous flight operations, maintenance and safety considerations, and personnel requirements related to the operation of multi-rotor \ac{suas} within the \projectmars{} inventory.
\\
\\
\noindent \textbf{Note:} This \ac{sop} must be present in the operational area of use. A copy will also be maintained in the \ac{kach} \projectmars{} office.

\section{Applicability}
        \subsubsection{Personnel}This \ac{sop} applies to all government and contractor personnel who are actively involved in air or ground operations related to \projectmars{}.  Operational procedures described in this \ac{sop} will be used to maintain prudent, safe operating practices and to ensure that appropriate response actions are taken in the event of an emergency.

        \subsubsection{Airworthiness} Airworthiness oversight of the \projectmars{} assets is documents via Level 3 \ac{awr} as issued by the Special Projects Team, \ac{uas} Division, \ac{srd} of RDECOM.  This \ac{sop} is suborindate to the Level 3 \ac{awr}.  Additionally, this \ac{sop} shall be subordinate to the \ac{sop} or regulations of the \ac{nas} controlled by teh \ac{faa}, restricted airspace, and any relevant agreements between the \ac{dod} and \ac{faa} where \projectmars{} is operating.

\section{Proponent and Authority}
        \subsubsection{Proponent}The proponent for this \ac{sop} is the \ac{kach} Project \ac{oic}.
\\
        \subsubsection{Exception/Waiver Authority}The proponent may approve written exceptions/waivers to this \ac{sop} that are consistent with controlling law, regulation, and safety policy. The proponent may delegate approval authority in writing to the Project \ac{ncoic}.
\\
        \subsubsection{Limits to Proponent Authority}Any change that alters \ac{opord}-controlled measures or authorities (for example, \ac{acm}s, altitude blocks, flight corridors/routes/boxes, \ac{tfb} approval lanes, or delegated command authority) requires action by the \ac{usma} G-3 via \ac{frago}.
\\
        \subsubsection{Recommended Changes}Users submit recommended changes via email to the Project \ac{ncoic} or Project \ac{oic}; the Project \ac{oic} adjudicates and publishes approved updates.
\\
        \subsubsection{Precedence}In the event of a conflict between this \ac{sop} and controlling law, regulation, or policy, the controlling document takes precedence. The proponent will correct the \ac{sop} at the next revision.
